Impact of Roe v. Wade’s Overturn: Abortion Restrictions and Infant Mortality in the U.S. Using a Difference-in-Differences Analysis

### 2024-12-02 paper/ExaminingtheimpactOverturnofRoev.Wade:BanningofAbortiononInfantMortalityRatesintheUnitedStatesUsingaDifference-in-DifferencesApproach.qmd
update paper with citation error

We conduct our infant death rate observational data analysis using the R
programming language [@citeR]. Our dataset, obtained from the CDC
Wonder[@cdc_wonder_2024], based on data as of December 2022, provides detailed state-level information on infant death rates in the United States, serving as the foundation for our study of the impact of abortion injunctions. CDC WONDER
(Wide-ranging ONline Data for Epidemiologic Research) is an online
system providing public health professionals and the public with access
to a wide range of the Centers for Disease Control and Prevention (CDC)
epidemiologic data and resources. This dataset offers detailed state-level information on infant death rates in the United States, serving as the foundation for our study of the impact of abortion injunctions.
Following the guidelines outlined in
@tellingstories, we analyze the impact of abortion injunctions on infant
death rates using the Difference-in-Difference method. This approach, shown in @sec-model, utilizes the main effects and the interaction between the legality of abortion in the
mother’s state of residence and the timing of the infant’s death as the predictor variable. Additionally, we include control variables, infant year of death, race, and age to account for potential confounding factors, and the mother’s state of residence as a random variable.

@@

Define $y_i$ as as the infant death rate per 1,000 live births, and
$\gamma_j$ as the random state effect for state j. 
$$
\begin{aligned}
y_i|\mu_i,\sigma &\sim \mbox{Normal}(\mu_i, \sigma)\\
\text{First Model}: \mu_i &= \beta_0 + \beta_1\cdot\text{After Injunction} + \beta_2\cdot\text{Abortion Illegal}\\&+\beta_3\cdot\text{After Injunction}\cdot\text{Abortion Illegal}_i+\gamma_j\\
\text{Second Model}:\mu_i &= \beta_0 + \beta_1\cdot\text{After Injunction}_i + \beta_2\cdot\text{Abortion Illegal}_i\\&+ \beta_3\cdot\text{After Injunction}_i\cdot\text{Abortion Illegal}_i\\&+\beta_4\cdot\text{Year of Death}_i+\beta_5\cdot\text{Age of Mother}_i+\beta_6\cdot\text{Mother's Single Race}_i+\gamma_j\\
\beta_0 &\sim \mbox{Normal}(0, 2.5)\\
\beta_1 &\sim \mbox{Normal}(0, 2.5)\\
\beta_2 &\sim \mbox{Normal}(0, 2.5)\\
\beta_3 &\sim \mbox{Normal}(0, 2.5)\\
\beta_4 &\sim \mbox{Normal}(0, 2.5)\\
\beta_5 &\sim \mbox{Normal}(0, 2.5)\\
\beta_6 &\sim \mbox{Normal}(0, 2.5)\\
\gamma_j&\sim \text{Normal}(0, \sigma_j^2)\\
\sigma &\sim \mbox{Exponential}(1)
\end{aligned}
$$

@@

Credibility intervals are the Bayesian equivalent of confidence
intervals, offering an understanding of the range of probable values for
a parameter @tellingstories. Using Bayesian estimation, we derive a
posterior distribution for each model coefficient. @fig-area-1 presents
the 95% credibility intervals for coefficients in the first model, while
@fig-area-2 illustrates the same for the second model. These intervals
represent the range within which 95% of the posterior probability mass
falls, offering a robust understanding of parameter uncertainty and the
influence of predictors under the Bayesian framework @tellingstories.

@@

Restrictive abortion laws can heighten maternal stress, anxiety, or
depression—particularly in unintended or high-risk
pregnancies—compromising maternal health and prenatal care. Research
@crugnola2022 highlights maternal mental health as a key determinant of
infant health outcomes, with conditions like chronic stress and depression adversely affecting prenatal behaviors, healthcare access,
and overall well-being. These compounded challenges, driven by societal and systemic barriers, can significantly increase risks of adverse
infant outcomes, including mortality.

>>>>

We conduct our infant death rate observational data analysis using the R
programming language [@citeR]. Our dataset, obtained from the CDC
Wonder[@cdc_wonder_2024], based on data as of December 2022, provides detailed state-level information on infant death rates in the United States, serving as the foundation for our study of the impact of abortion injunctions. CDC WONDER
(Wide-ranging ONline Data for Epidemiologic Research) is an online
system providing public health professionals and the public with access
to a wide range of the Centers for Disease Control and Prevention (CDC)
epidemiologic data and resources. This dataset offers detailed state-level information on infant death rates in the United States, serving as the foundation for our study of the impact of abortion injunctions.
Following the guidelines outlined in
@tellingstories, we analyze the impact of abortion injunctions on infant
death rates using the Difference-in-Difference method. This approach, shown in @sec-model, utilizes the main effects and the interaction between the legality of abortion in the mother’s state of residence and the timing of the infant’s death as the predictor variable. Additionally, we include the infants' year of death, race, and age as control variables to account for potential confounding factors, and the mother’s state of residence as a random variable.

@@

Define $y_i$ as as the infant death rate per 1,000 live births, and
$\gamma_j$ as the random state effect for state $j$. 
$$
\begin{aligned}
y_i|\mu_i,\sigma &\sim \mbox{Normal}(\mu_i, \sigma)\\
\text{First Model}: \mu_i &= \beta_0 + \beta_1\cdot\text{After Injunction} + \beta_2\cdot\text{Abortion Illegal}\\&+\beta_3\cdot\text{After Injunction}\cdot\text{Abortion Illegal}_i+\gamma_j\\
\text{Second Model}:\mu_i &= \beta_0 + \beta_1\cdot\text{After Injunction}_i + \beta_2\cdot\text{Abortion Illegal}_i\\&+ \beta_3\cdot\text{After Injunction}_i\cdot\text{Abortion Illegal}_i\\&+\beta_4\cdot\text{Year of Death}_i+\beta_5\cdot\text{Age of Mother}_i+\beta_6\cdot\text{Mother's Single Race}_i+\gamma_j\\
\beta_0 &\sim \mbox{Normal}(0, 2.5)\\
\beta_1 &\sim \mbox{Normal}(0, 2.5)\\
\beta_2 &\sim \mbox{Normal}(0, 2.5)\\
\beta_3 &\sim \mbox{Normal}(0, 2.5)\\
\beta_4 &\sim \mbox{Normal}(0, 2.5)\\
\beta_5 &\sim \mbox{Normal}(0, 2.5)\\
\beta_6 &\sim \mbox{Normal}(0, 2.5)\\
\gamma_j&\sim \text{Normal}(0, \sigma_j^2)\\
\sigma &\sim \mbox{Exponential}(1)
\end{aligned}
$$

@@

Credibility intervals are the Bayesian equivalent of confidence
intervals, offering an understanding of the range of probable values for
a parameter [@tellingstories]. Using Bayesian estimation, we derive a
posterior distribution for each model coefficient. @fig-area-1 presents
the 95% credibility intervals for coefficients in the first model, while
@fig-area-2 illustrates the same for the second model. These intervals
represent the range within which 95% of the posterior probability mass
falls, offering a robust understanding of parameter uncertainty and the
influence of predictors under the Bayesian framework [@tellingstories].

@@

Restrictive abortion laws can heighten maternal stress, anxiety, or
depression—particularly in unintended or high-risk
pregnancies—compromising maternal health and prenatal care. Research from
@crugnola2022 highlights maternal mental health as a key determinant of
infant health outcomes, with conditions like chronic stress and depression adversely affecting prenatal behaviors, healthcare access,
and overall well-being. These compounded challenges, driven by societal and systemic barriers, can significantly increase risks of adverse
infant outcomes, including mortality.
